Building a continuous benchmarking ecosystem in bioinformatics

### 2025-02-12 original_submission/manifesto-main.tex
Add figure draft
Mark changes

\affil[*]{\{mark.robinson,izaskun.mallona\}@mls.uzh.ch}

@@

\hfill

@@

\section*{Funding}

MDR acknowledges funding from the Swiss National Science Foundation (grants 200021\_212940 and 310030\_204869) as well as support from swissuniversities P5 Phase B funding (project 23-36\_14). CS is supported by the Novartis Research Foundation.

\section*{Disclosure statement}

No potential conflict of interest is reported by the authors

>>>>

\affil[*]{\{izaskun.mallona,mark.robinson\}@mls.uzh.ch}

@@

%\hfill

@@

% the two sections below have been added to the online system now
\section*{Funding}
MDR acknowledges funding from the Swiss National Science Foundation (grants 200021\_212940 and 310030\_204869) as well as support from swissuniversities P5 Phase B funding (project 23-36\_14). CS is supported by the Novartis Research Foundation.

\section*{Competing interests}
No conflicts of interest are declared by the authors.

### 2025-02-12 review_cycle_1/highlighted_manifesto_main.tex
Prior to Mark's edits
Mark changes

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{figures/fig_benchmarking.pdf}
    {\color{red} \caption[Run types, properties, capabilities and features.]{Solo- and community-benchmarks, workflows, and benchmarking system-enabled (\textit{bs}) benchmarks differ in run properties and their consequences in terms of capabilities and features, including whether: inputs, truths, hidden truths, metrics or run results are available for readers; code is free software, software environments are managed, and execution automated; there is a clear track of (human) contributors; results are run in comparable hardware; an entrypoint to results or result interpretation are provided. These run intrinsic properties have consequences in terms of features (i.e. transparency, forkability, availability for metaanalysis) and technical capabilities (i.e. software versioning or ability to run anonymously and locally (stealth mode). \label{fig:definitions}}
    }
\end{figure}

@@

Benchmarking as a process deals with tasks other than workflow definition and execution, including: collecting and keeping track of contributions from users; provisioning appropriate hardware; handling a reproducible and efficient software stack {\color{red}(the advantages of decoupling environments handling from workflow execution is further discussed in section~`Reproducible software environments')}; managing storage and access-control lists; rendering results dashboards and summarizing results; versioning code, workflow runs, and files; and providing extensive documentation from defining scope or contribution guidelines, to transparent logging and crediting contributions and disseminating results {\color{red}(Figure~\ref{fig:definitions})}.

@@

{\color{red}
Aside of task completion-tailored metrics, computing performance (i.e. wallclock, on-disk read and write operations), requirements (e.g. GPUs, minimum required memory) and their scalability depending on the input size (e.g. whether runtimes increase linearly, quadratically, etc) are key, and routine, benchmarking metrics. To measure these accurately, benchmarking systems ideally would provide means to: i, characterize the hardware running the benchmark, so methods runtimes are comparable if run on homogeneous systems;  ii, track performances (CPU or GPU usage, memory) when running each benchmark component; iii, introduce little overhead during the profiling (so profiling does not considerably slow down runtimes); iv, collect variability in performance by running each method several times; v, go beyond reporting average resources usage and provide them over execution time (e.g. so embarrasingly parallel algorithms can be told apart from largely single-cored runs with few parallelism bursts).
}

@@

{\color{red}
Mature communities are key in sustaining and reflecting the existing challenge in a given field. For instance, CASP, a continuous benchmarking by nature, has a established community defining the critical assesments scope, input data, target definitions, and composite and normalized metrics to equally weight multiple performance scores \cite{simpkin2023tertiary}. At the continuous benchmark level, CASP keeps track of past performances and introduces metrics to evaluate possible artifacts, i.e. whether round progress could be arising from different target difficulty \cite{kryshtafovych2019critical}. To facilitate community contribution, we believe benchmarking systems can lower the barrier to contributing, hence making the contribution efforts less likely to be biased. 
}

@@

{\color{red} 
The responsibility of handling the software environment poses an interesting challenge to benchmarking systems. On one hand, some workflow managers provide means to install and activate software and re-run workflows when software specification changes, even if the workflow topology remains the same. For these, software is the workflow manager's responsability. On the other hand, some benchmarks might be addressed to comparing the same workflow topology run with different software backends (i.e., with conda vs compilled) or with different versions of some software dependencies components (i.e. several samtools/htslib releases). Benchmarking systems should provide enough flexibility to take control and make the latter possible.

Finally, scientific software management handling is known for its nuances, including abundant dependencies with known incompatibilities and complex installation recipes \cite{droge2023-ax}. Scientific software management is handled via traditional (linux) package managers (Ri.e. PMs, DEBs), central software stacks and enviornment modules (i.e. Lmod), conda, containers, or read-only network filesystems providing ready-to-use installations (CernVM-FS). These strategies are under active development: EESSI \cite{droge2023-ax} and the Digital Research Alliance of Canada \cite{Boissonneault2019-wm} are effectively providing multi-purpose, reusable and efficient software installations via CernVM-FS; conda is speeding up their deployment procedures (via py-rattler/pixi) and allowing compilation to Web assembly (to execute code within Web browser)s; or apptainer (former Singularity) provides rootless containarization suitable to run in scientific clusters. 
}

@@

Even a simple, single-person benchmark comparing methods relies on multiple contributions, from dataset creators and method developers to system software and workflow authors. Hence, data, code and operating system dependencies are all accessible during the benchmarking process. For reproducibility, transparency and to build trust \cite{Laine2007-py}, this accessibility and freedom should be extended to others to run (re-run), copy (including fork), distribute (e.g., snapshot), study (e.g., read how someone is implementing a method), change (bug fix, contribute) and improve (extend) benchmarks. Such freedoms are granted (or restricted) by software and content licenses, as are copyright and authorship and attribution \cite{Kreutzer2014-ua}. Without explicit licenses, even publicly available code is copyrighted by default, and reuse is restricted (to reproduce, distribute, or create derivative works) \cite{Kreutzer2014-ua}. A concise summary of license rights and restrictions, and compatibility across commonly-used open-source licenses can be found in Table~\ref{tab:licenses}. Benchmark contributions should be licensed and associated to scholarly names including email and ORCID (as the canonical identifier), hence allowing seamless reuse, transparency and attribution. {\color{red} As a consequence, properly licensed benchmarking pieces (e.g. a procedure to compute a clustering metric; or a clustering method) could be reused across benchmarks; we envision assisted benchmark design tools to leverage and reuse benchmark components, perhaps guided by AI.}

@@

{\color{red}
\subsection*{Funding}
@Mark
}
\section*{Author contributions}

>>>>

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{figures/fig_benchmarking.pdf}
    {\color{red} \caption[Run types, properties, capabilities and features.]{Solo and community benchmarks, workflows, and benchmarking-system-enabled (\textit{bs}) benchmarks all differ in run properties and their consequences in terms of capabilities and features, including whether: inputs, truths, hidden truths, metrics or run results are available for readers; code is free software, software environments are managed, and execution automated; there is a clear track of (human) contributors; results are run in comparable hardware; an entrypoint to results or result interpretation are provided. These run-intrinsic properties have consequences in terms of features (i.e., transparency, forkability, availability for metaanalysis) and technical capabilities (i.e. software versioning or ability to run anonymously and locally in stealth mode). \label{fig:definitions}}
    }
\end{figure}

@@

Benchmarking as a process deals with tasks other than workflow definition and execution, including: collecting and keeping track of contributions from users; provisioning appropriate hardware; handling a reproducible and efficient software stack {\color{red}(the advantages of decoupling environment handling from workflow execution is further discussed in section~`Reproducible software environments')}; managing storage and access-control lists; rendering results dashboards and summarizing results; versioning code, workflow runs, and files; and providing extensive documentation from defining scope or contribution guidelines, to transparent logging and crediting contributions and disseminating results {\color{red}(Figure~\ref{fig:definitions})}.

@@

{\color{red}
Aside of task completion-tailored metrics, computing performance (i.e., wallclock, on-disk read-write operations), requirements (e.g., GPUs, minimum memory) and their scalability depending on the input size (e.g., whether runtimes increase linearly, quadratically, etc) are key routine benchmarking metrics. To measure these accurately, benchmarking systems ideally would provide means to: i) characterize the hardware running the benchmark, so method runtimes are comparable if run on homogeneous systems;  ii) track performance (CPU or GPU usage, memory) when running each benchmark component; iii) introduce minimal overhead during profiling (so profiling does not considerably slow down runtimes); iv) collect variability in performance by running each method several times; and, v) go beyond reporting average resource usage and provide them over execution time (e.g., so embarrassingly-parallel algorithms can be differentiated from largely single-cored runs with few parallelism bursts).
}

@@

{\color{red}
Mature communities are key to sustaining and reflecting the existing challenges in a given field. For instance, CASP, a continuous benchmark by nature, has a established community that defines the scope of critical assessment, input data, target definitions, and composite and normalized metrics to equally weight multiple performance scores \cite{simpkin2023tertiary}. At the continuous benchmark level, CASP keeps track of past performances and introduces metrics to evaluate possible artifacts, i.e., whether round progress could arises from different target difficulty \cite{kryshtafovych2019critical}. To facilitate community contribution, we believe that benchmarking systems can lower the barrier to contributing, hence making the contribution efforts less likely to be biased. 
}

@@

{\color{red} 
The responsibility of handling software environments poses an interesting challenge to benchmarking systems. On one hand, some workflow managers provide means to install and activate software and re-run workflows when software specification changes, even if the workflow topology remains the same. For these, software is the workflow manager's responsibility. On the other hand, some benchmarks might be explicitly designed to compare the same workflow topology run with different software backends (i.e., with conda vs compiled) or with different versions of a dependency (e.g., samtools/htslib releases). Benchmarking systems should provide enough flexibility to take control and make the latter possible.

Finally, scientific software management handling is known for its nuances, including abundant dependencies with known incompatibilities and complex installation recipes \cite{droge2023-ax}. Scientific software management is handled via traditional (linux) package managers (RPMs, DEBs), central software stacks and environment modules (i.e., Lmod), conda, containers, or read-only network filesystems providing ready-to-use installations (CernVM-FS). These strategies are under active development: EESSI \cite{droge2023-ax} and the Digital Research Alliance of Canada \cite{Boissonneault2019-wm} are effectively providing multi-purpose, reusable and efficient software installations via CernVM-FS; conda is speeding up their deployment procedures (via py-rattler/pixi) and allowing compilation to Web assembly (to execute code within Web browser); or apptainer (former Singularity) provides rootless containerization suitable to run in scientific clusters. 
}

@@

Even a simple, single-person benchmark comparing methods relies on multiple contributions, from dataset creators and method developers to system software and workflow authors. Hence, data, code and operating system dependencies are all accessible during the benchmarking process. For reproducibility, transparency and to build trust \cite{Laine2007-py}, this accessibility and freedom should be extended to others to run (re-run), copy (including fork), distribute (e.g., snapshot), study (e.g., read how someone is implementing a method), change (bug fix, contribute) and improve (extend) benchmarks. Such freedoms are granted (or restricted) by software and content licenses, as are copyright and authorship and attribution \cite{Kreutzer2014-ua}. Without explicit licenses, even publicly available code is copyrighted by default, and reuse is restricted (to reproduce, distribute, or create derivative works) \cite{Kreutzer2014-ua}. A concise summary of license rights and restrictions, and compatibility across commonly-used open-source licenses can be found in Table~\ref{tab:licenses}. Benchmark contributions should be licensed and associated to scholarly names including email and ORCID (as the canonical identifier), hence allowing seamless reuse, transparency and attribution. {\color{red} As a consequence, properly-licensed benchmark components (e.g., a procedure to compute a clustering metric; or a clustering method) could be reused across benchmarks; we envision assisted benchmark design tools to leverage and reuse benchmark components, perhaps guided by AI.}

@@

{\color{red}
\section*{Funding}
MDR acknowledges funding from the Swiss National Science Foundation (grants 200021\_212940 and 310030\_204869) as well as support from swissuniversities P5 Phase B funding (project 23-36\_14). CS is supported by the Novartis Research Foundation.

\section*{Conflict of interest}

We report no conflict of interest.
}
\section*{Author contributions}
